---
title: "synthesis code"
author: "Patrick Das"
date: "2025-07-07"
output:
  word_document: default
  html_document:
    df_print: paged
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data cleaning and transformation

## Load packages

```{r error=FALSE, message=FALSE, warning=FALSE, tidy=TRUE}
library(tidyr)
library(readr)
library(dplyr)
library(readr)
library(ggplot2)
library(ggrepel)
library(nnet)
library(lmtest)
library(nortest)
library(spdep)
library(lme4)
library(lmerTest) 
library(MuMIn)

data <- read_tsv("synthesis aggregrate data.txt")

data <- data[, colSums(is.na(data)) == 0] #dropping columns with NA values

data[data == "N/A"] <- NA #rewriting "NA" values in text to reflect actual NA

data <- data %>% drop_na() #dropping observations with NA values 


```

## Data preparation

### Converting variables to factors

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
data <- data %>%
  mutate(across(where(is.character), as.factor)) #converting chr to factors

```

### Cleaning column names

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Create clean, consistent column names
clean_names <- names(data)

# Replace spaces and punctuation with underscores
clean_names <- gsub(" ", "_", clean_names)             # spaces → _
clean_names <- gsub("\\?", "", clean_names)            # remove question marks
clean_names <- gsub("\\(", "", clean_names)            # remove (
clean_names <- gsub("\\)", "", clean_names)            # remove )
clean_names <- gsub("\\/", "_", clean_names)           # slashes → _
clean_names <- gsub("\\+", "plus", clean_names)        # plus signs → 'plus'
clean_names <- gsub("-", "_", clean_names)             # hyphens → _
clean_names <- gsub("__+", "_", clean_names)           # collapse double underscores
clean_names <- gsub("[^[:alnum:]_]", "", clean_names)  # remove anything not alphanumeric or underscore
clean_names <- tolower(clean_names)                    # make all lowercase (optional)

# Apply new names to the dataframe
names(data) <- clean_names

# Applying custom mapping
names(data)[names(data) == "granular_case_marking_alignment_type"] <- "granular_alignment"
names(data)[names(data) == "case_marking_complexity_distinctions_plus_optionality"] <- "case_marking_complexity"
names(data)[names(data) == "spoken_as_an_l2"] <- "L2"
names(data)[names(data) == "speaker_population"] <- "population"
names(data)[names(data) == "agriculture_intensity_based_on_ea028"] <- "agricultural_intensity"
names(data)[names(data) == "political_organization_ea033"] <- "political_organization"
names(data)[names(data) == "altitude_sampled_raster_value_from_dem"] <- "altitude"
names(data)[names(data) == "subfamily_node_below_top_level"] <- "subfamily"
names(data)[names(data) == "language_family_top_level_node"] <- "family"

```

### Creating/transforming variables

#### Binarizing agricultural intensity
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
data$agriculture_binary <- ifelse(
  data$agricultural_intensity == "Intensive/irrigated", 1, 0
)
table(data$agriculture_binary)
```
#### Binarizing political organization

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
data$political_organization_binary <- ifelse(
  data$political_organization == "State", 1, 0
  )
table(data$political_organization_binary)

```
#### Binarizing hill/valley (needed for ecological mode)

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Create hill_binary variable (1 = Hill, 0 = Valley or Split)
data$hill_binary <- ifelse(data$hill_valley == "Hill", 1, 0)

table(data$hill_valley, data$hill_binary)
```

#### Scaling variables

Scaling population

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#let's scale populaiton
data$scaled_population <- scale(data$population)
# Scaling geographic predictors
data$scaled_altitude <- scale(data$altitude)[, 1]
data$scaled_stdev_slope <- scale(data$stdev_of_slope)[, 1]
```

Scaling complexity scores

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#First, individual measures of complexity are scaled
data$scaled_case_marking_complexity <- scale(data$case_marking_complexity)
data$scaled_form_complexity <- scale(data$form_complexity)
data$scaled_cell_complexity <- scale(data$cell_complexity)
```

# Initial Analysis (with Limbu)

## Hill/Valley classification 
Predicting the complexity of a language system based on hill/valley classification


```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Cell complexity model
model_cell <- lmer(scaled_cell_complexity ~ hill_valley + (1 | subfamily), data = data)
summary(model_cell)

# Form complexity model
model_form <- lmer(scaled_form_complexity ~ hill_valley + (1 | subfamily), data = data)
summary(model_form)

# Case marking complexity model
model_case <- lmer(scaled_case_marking_complexity ~ hill_valley + (1 | subfamily), data = data)
summary(model_case)


```
# Visualizing data, identifying and removing outliers, and testing assumptions

## Visualizing complexity against hill/valley via box-plots
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Reshape the data
data_long <- data %>%
  select(hill_valley, scaled_cell_complexity, scaled_form_complexity, scaled_case_marking_complexity) %>%
  pivot_longer(
    cols = starts_with("scaled_"),
    names_to = "complexity_type",
    values_to = "score"
  ) %>%
  mutate(complexity_type = case_when(
    complexity_type == "scaled_cell_complexity" ~ "Cell Complexity",
    complexity_type == "scaled_form_complexity" ~ "Form Complexity",
    complexity_type == "scaled_case_marking_complexity" ~ "Case Marking Complexity"
  ))
ggplot(data_long, aes(x = hill_valley, y = score, fill = hill_valley)) +
  geom_boxplot() +
  facet_wrap(~ complexity_type, scales = "free_y", 
             labeller = label_wrap_gen(width = 15)) +
  labs(
    title = "Morphological Complexity by Hill/Valley",
    x = "Societal Setting", y = "Scaled Complexity Score"
  ) +
  scale_fill_brewer(palette = "Pastel2") +
  theme_minimal() +
  theme(legend.position = "none")

```


## sensitivity testing

Identifying outliers and influential values
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
form_cooks <- cooks.distance(model_form)
cell_cooks <- cooks.distance(model_cell)
case_cooks <- cooks.distance(model_case)

# Set threshold
threshold <- 4 / nrow(data)

# Identify outliers
form_outlier <- which(form_cooks > threshold)
cell_outlier <- which(cell_cooks > threshold)
case_outlier <- which(case_cooks >threshold)


#and most influential
which.max(form_cooks)
which.max(cell_cooks)
which.max(case_cooks)

# For Limbu
limbu_form <- form_cooks[data$language == "Limbu"]
limbu_cell <- cell_cooks[data$language == "Limbu"]

# For Bori-Karko
bori_case <- case_cooks[data$language == "Bori-Karko"]

#printing out most influential observations
# For form complexity
data[which.max(form_cooks), ]
limbu_form
# For cell complexity
data[which.max(cell_cooks), ]
limbu_cell
# For case complexity
data[which.max(case_cooks), ]
bori_case
```
## Visualizing outliers

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Creating tidy Cook's data with language preserved
cook_data <- data %>%
  mutate(
    obs = row_number(),
    form = form_cooks,
    cell = cell_cooks,
    case = case_cooks
  ) %>%
  select(obs, language, form, cell, case)

# Pivoting to long format for ggplot
cook_long <- cook_data %>%
  pivot_longer(cols = c(form, cell, case),
               names_to = "model", values_to = "cooks") %>%
  mutate(
    outlier = cooks > threshold,
    label = ifelse(outlier, as.character(language), NA)
  )

# Plotting with labeled outliers
ggplot(cook_long, aes(x = obs, y = cooks, color = outlier, label = label)) +
  geom_point() +
  geom_hline(yintercept = threshold, linetype = "dashed", color = "red") +
  geom_text(nudge_y = 0.01, check_overlap = TRUE, size = 3.2) +
  facet_wrap(~ model, scales = "free_y") +
  theme_minimal() +
  labs(title = "Cook's Distance with Influential Languages Labeled",
       x = "Observation", y = "Cook's Distance") +
  scale_color_manual(values = c("black", "firebrick"))+
  theme(plot.margin = margin(10, 20, 10, 10))  # top, right, bottom, left (in pts)

```

## Creating a new dataset without Limbu and Bodo-Mech

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
data_sensitivity <- data[data$language != "Limbu", ]

```

## Visualizing variables

### Visualizing ecological variables against hill/valley classification

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Create long-format data for both terrain variables
terrain_data <- data %>%
  select(hill_valley, altitude, stdev_of_slope) %>%
  pivot_longer(
    cols = c(altitude, stdev_of_slope),
    names_to = "terrain_variable",
    values_to = "value"
  ) %>%
  mutate(terrain_variable = case_when(
    terrain_variable == "altitude" ~ "Elevation (meters)",
    terrain_variable == "stdev_of_slope" ~ "Terrain roughness"
  ))

# Create faceted boxplot
ggplot(terrain_data, aes(x = hill_valley, y = value, fill = hill_valley)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~ terrain_variable, scales = "free_y") +
  labs(
    title = "Proposed Ecological Variables vs. Hill/Valley Classification",
    x = "Societal Classification", 
    y = "Ecological variables",
    fill = "Classification"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12)
  )
```


# Analysis (after removing Limbu)

## Predicting complexity based on hill/valley classification 

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Cell complexity model
model_cell_sensitivity <- lmer(scaled_cell_complexity ~ hill_valley + (1 | subfamily),
                               data = data_sensitivity)

# Form complexity model
model_form_sensitivity <- lmer(scaled_form_complexity ~ hill_valley + (1 | subfamily),
                                data = data_sensitivity)

# Case complexity model
model_case_sensitivity <- lmer(scaled_case_marking_complexity ~ hill_valley + (1 | subfamily),
                                data = data_sensitivity)

# View summaries
summary(model_cell_sensitivity)
summary(model_form_sensitivity)
summary(model_case_sensitivity)

```
No significant effect (but directionality changed for Cell/Form models)

## Testing assumptions

### Plotting residuals

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
plot_residuals <- function(model, title) {
  # Create plot
  plot(fitted(model), residuals(model),
       xlab = "Fitted Values", ylab = "Residuals",
       main = paste("Residuals vs Fitted (", title, ")", sep=""))
  # Add reference line
  abline(h = 0, col = "red", lty = 2)
}
par(mfrow = c(2, 2))  # Set up a 2x2 plotting area

# Plot each model
plot_residuals(model_cell_sensitivity, "Cell Complexity")
plot_residuals(model_form_sensitivity, "Form Complexity")
plot_residuals(model_case_sensitivity, "Case Marking Complexity")

```

### Q-Q line and Shapiro wilk test

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Q-Q plot cell
qqnorm(resid(model_cell_sensitivity), main = "Q-Q Plot (Cell Complexity)")
qqline(resid(model_cell_sensitivity), col = "red")

# Shapiro-Wilk test
shapiro.test(resid(model_cell_sensitivity))

# Q-Q plot
qqnorm(resid(model_form_sensitivity), main = "Q-Q Plot (Form Complexity)")
qqline(resid(model_form_sensitivity), col = "red")

# Shapiro-Wilk test
shapiro.test(resid(model_form_sensitivity))

# Q-Q plot
qqnorm(resid(model_case_sensitivity), main = "Q-Q Plot (Case Marking Complexity)")
qqline(resid(model_case_sensitivity), col = "red")

# Shapiro-Wilk test
shapiro.test(resid(model_case_sensitivity))
```

### Normality of random effects

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Extract random effects
ranef_subfam <- ranef(model_cell_sensitivity)$subfamily

# Check normality of random effects
qqnorm(ranef_subfam[,1])
qqline(ranef_subfam[,1])
shapiro.test(ranef_subfam[,1])
```

## Predicting Hill/Valley classification from ecological variables

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Binarizing hill/valley classification: Hill = 1, others = 0
data$hill_binary <- ifelse(data$hill_valley == "Hill", 1, 0)

# Fitting logistic regression model
hill_ecology_model <- glm(
  hill_binary ~ scaled_altitude + scaled_stdev_slope,
  family = binomial,
  data = data_sensitivity
)

hill_altitude <- glm(hill_binary ~ scaled_altitude, family = binomial, data = data_sensitivity)
hill_slope <- glm(hill_binary ~ scaled_stdev_slope, family = binomial, data = data_sensitivity)

# Summarizing model output
summary(hill_ecology_model)
summary(hill_altitude)
summary(hill_slope)


```
## Analysing complexity as predicted by sociocultural variables

### Complexity as predicted by L2_status

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# 1. L2 STATUS EFFECTS
# Cell complexity model
cell_L2_mixed <- lmer(scaled_cell_complexity ~ L2 + (1|subfamily), data = data_sensitivity)

# Form complexity model
form_L2_mixed <- lmer(scaled_form_complexity ~ L2 + (1|subfamily), data = data_sensitivity)

# Case marking complexity model
case_L2_mixed <- lmer(scaled_case_marking_complexity ~ L2 + (1|subfamily), data = data_sensitivity)

# View L2 model summaries
summary(cell_L2_mixed)
summary(form_L2_mixed)
summary(case_L2_mixed)
```

#### If Limbu is added back in

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# 1. L2 STATUS EFFECTS
# Cell complexity model
cell_L2_mixed_orig <- lmer(scaled_cell_complexity ~ L2 + (1|subfamily), data = data)

# Form complexity model
form_L2_mixed_orig <- lmer(scaled_form_complexity ~ L2 + (1|subfamily), data = data)

# Case marking complexity model
case_L2_mixed_orig <- lmer(scaled_case_marking_complexity ~ L2 + (1|subfamily), data = data)

# View L2 model summaries
summary(cell_L2_mixed_orig)
summary(form_L2_mixed_orig)
summary(case_L2_mixed_orig)
```

#### Is effect of L2 on cell_complexity robust to dropping languages?


```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Create a dataframe to store results
drop_one_results <- data.frame(
  dropped_language = character(),
  estimate = numeric(),
  std_error = numeric(),
  p_value = numeric(),
  significant = logical(),
  stringsAsFactors = FALSE
)

# Loop through each language in data_sensitivity
for (lang in unique(data_sensitivity$language)) {
  # Create dataset with one additional language dropped
  temp_data <- data_sensitivity[data_sensitivity$language != lang, ]
  
  # Fit the model
  temp_model <- lmer(scaled_cell_complexity ~ L2 + (1|subfamily), data = temp_data)
  
  # Extract model summary
  model_summary <- summary(temp_model)
  
  # Get L2 coefficient (second row in fixed effects table)
  l2_coef <- model_summary$coefficients[2, "Estimate"]
  l2_se <- model_summary$coefficients[2, "Std. Error"]
  l2_p <- model_summary$coefficients[2, "Pr(>|t|)"]
  
  # Add to results dataframe
  drop_one_results <- rbind(drop_one_results, data.frame(
    dropped_language = lang,
    estimate = l2_coef,
    std_error = l2_se,
    p_value = l2_p,
    significant = l2_p < 0.05
  ))
}

# Sort results by p-value
drop_one_results <- drop_one_results[order(drop_one_results$p_value), ]

# Look at summary statistics
cat("Number of iterations where L2 effect is significant:", sum(drop_one_results$significant), 
    "out of", nrow(drop_one_results), "\n")
cat("Range of L2 coefficient estimates:", 
    round(min(drop_one_results$estimate), 3), "to", 
    round(max(drop_one_results$estimate), 3), "\n")
cat("Median p-value:", round(median(drop_one_results$p_value), 3), "\n")

# Plot the results
ggplot(drop_one_results, aes(x = reorder(dropped_language, p_value), y = estimate, 
                           fill = significant)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = estimate - std_error, ymax = estimate + std_error), width = 0.2) +
  coord_flip() +
  labs(title = "L2 Effect on Cell Complexity - Drop-One Analysis",
       subtitle = "Effect of removing each language after Limbu already removed",
       x = "Second Removed Language", 
       y = "L2 Coefficient Estimate",
       fill = "Significant (p < 0.05)") +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkgray")
```
### Complexity ~ Population
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# 2. POPULATION SIZE EFFECTS
cell_pop_mixed <- lmer(scaled_cell_complexity ~ scaled_population + (1|subfamily), data = data_sensitivity)
form_pop_mixed <- lmer(scaled_form_complexity ~ scaled_population + (1|subfamily), data = data_sensitivity)
case_pop_mixed <- lmer(scaled_case_marking_complexity ~ scaled_population + (1|subfamily), data = data_sensitivity)

# View population model summaries
summary(cell_pop_mixed)
summary(form_pop_mixed)
summary(case_pop_mixed)
```

### Complexity ~ agricultural intensity
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}


# 3. AGRICULTURAL INTENSITY EFFECTS
cell_agr_mixed <- lmer(scaled_cell_complexity ~ agriculture_binary + (1|subfamily), data = data_sensitivity)
form_agr_mixed <- lmer(scaled_form_complexity ~ agriculture_binary + (1|subfamily), data = data_sensitivity)
case_agr_mixed <- lmer(scaled_case_marking_complexity ~ agriculture_binary + (1|subfamily), data = data_sensitivity)

# View agriculture model summaries
summary(cell_agr_mixed)
summary(form_agr_mixed)
summary(case_agr_mixed)
```
### complexity ~ political organization
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# 4. POLITICAL ORGANIZATION EFFECTS
cell_pol_mixed <- lmer(scaled_cell_complexity ~ political_organization_binary + (1|subfamily), data = data_sensitivity)
form_pol_mixed <- lmer(scaled_form_complexity ~ political_organization_binary + (1|subfamily), data = data_sensitivity)
case_pol_mixed <- lmer(scaled_case_marking_complexity ~ political_organization_binary + (1|subfamily), data = data_sensitivity)

# View political organization model summaries
summary(cell_pol_mixed)
summary(form_pol_mixed)
summary(case_pol_mixed)
```

## R-squared values for major analyses

### Complexity ~ Hill/Valley models

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
r.squaredGLMM(model_case_sensitivity)
r.squaredGLMM(model_cell_sensitivity)
r.squaredGLMM(model_form_sensitivity)
```
### Hill/Valley ~ ecology models

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
r.squaredGLMM(hill_ecology_model)
r.squaredGLMM(hill_altitude)
r.squaredGLMM(hill_slope)
```
### Complexity ~ sociocultural variables

#### L2_status
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
r.squaredGLMM(cell_L2_mixed)
r.squaredGLMM(form_L2_mixed)
r.squaredGLMM(case_L2_mixed)
```
#### population size
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
r.squaredGLMM(cell_pop_mixed)
r.squaredGLMM(form_pop_mixed)
r.squaredGLMM(case_pop_mixed)
```
#### agricultural_intensity
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
r.squaredGLMM(cell_agr_mixed)
r.squaredGLMM(form_agr_mixed)
r.squaredGLMM(case_agr_mixed)
```
#### political_organization
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
r.squaredGLMM(cell_pol_mixed)
r.squaredGLMM(form_pol_mixed)
r.squaredGLMM(case_pol_mixed)
```

# Additional analyses

## Ecological variables as a predictor of complexity


```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
## ALTITUDE MODELS
# Altitude models - full dataset
cell_alt_mixed <- lmer(scaled_cell_complexity ~ scaled_altitude + (1|subfamily), data = data)
form_alt_mixed <- lmer(scaled_form_complexity ~ scaled_altitude + (1|subfamily), data = data)
case_alt_mixed <- lmer(scaled_case_marking_complexity ~ scaled_altitude + (1|subfamily), data = data)

# Altitude models - sensitivity dataset (without Limbu)
cell_alt_mixed_sens <- lmer(scaled_cell_complexity ~ scaled_altitude + (1|subfamily), data = data_sensitivity)
form_alt_mixed_sens <- lmer(scaled_form_complexity ~ scaled_altitude + (1|subfamily), data = data_sensitivity)
case_alt_mixed_sens <- lmer(scaled_case_marking_complexity ~ scaled_altitude + (1|subfamily), data = data_sensitivity)

# Slope variability models - full dataset
cell_slope_mixed <- lmer(scaled_cell_complexity ~ scaled_stdev_slope + (1|subfamily), data = data)
form_slope_mixed <- lmer(scaled_form_complexity ~ scaled_stdev_slope + (1|subfamily), data = data)
case_slope_mixed <- lmer(scaled_case_marking_complexity ~ scaled_stdev_slope + (1|subfamily), data = data)

# Slope variability models - sensitivity dataset (without Limbu)
cell_slope_mixed_sens <- lmer(scaled_cell_complexity ~ scaled_stdev_slope + (1|subfamily), data = data_sensitivity)
form_slope_mixed_sens <- lmer(scaled_form_complexity ~ scaled_stdev_slope + (1|subfamily), data = data_sensitivity)
case_slope_mixed_sens <- lmer(scaled_case_marking_complexity ~ scaled_stdev_slope + (1|subfamily), data = data_sensitivity)

# Get summaries for altitude models
summary(cell_alt_mixed)
summary(form_alt_mixed)
summary(case_alt_mixed)
summary(cell_alt_mixed_sens)
summary(form_alt_mixed_sens)
summary(case_alt_mixed_sens)

# Get summaries for slope variability models
summary(cell_slope_mixed)
summary(form_slope_mixed)
summary(case_slope_mixed)
summary(cell_slope_mixed_sens)
summary(form_slope_mixed_sens)
summary(case_slope_mixed_sens)

```

## Languages miscategorized as Hill/Valley

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Check if there are Hill languages with below-average cell complexity
hill_languages <- filter(data, hill_valley == "Hill") %>%
  mutate(below_average = scaled_cell_complexity < 0) %>%
  arrange(scaled_cell_complexity)

print("Hill languages sorted by cell complexity (lowest first):")
print(select(hill_languages, language, scaled_cell_complexity, below_average))
```

### Plotting

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Create classification based on theoretical expectations using mean (0)
data_with_theoretical_mismatches <- data %>%
  mutate(
    # Classify based on complexity relative to the mean (0)
    theoretical_status = case_when(
      # Hill languages with below-mean complexity
      hill_valley == "Hill" & scaled_cell_complexity < 0 ~ 
        "Misclassified (Low Complexity Hill)",
      
      # Valley languages with above-mean complexity
      hill_valley == "Valley" & scaled_cell_complexity > 0 ~ 
        "Misclassified (High Complexity Valley)",
      
      # Everything else matches theoretical expectations
      TRUE ~ "Matches Theory"
    )
  )

# Check the distribution of theoretical status
print(table(data_with_theoretical_mismatches$theoretical_status))

# Create scatter plot
ggplot(data_with_theoretical_mismatches, aes(x = scaled_cell_complexity, y = scaled_case_marking_complexity)) +
  # Adding quadrant lines at means (0 for scaled variables)
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  
  # Plot points
  geom_point(aes(
    shape = hill_valley,
    color = theoretical_status,
    size = theoretical_status
  )) +
  
  # Add labels for misclassified languages
  geom_text_repel(
    data = filter(data_with_theoretical_mismatches, theoretical_status != "Matches Theory"),
    aes(label = language),
    box.padding = 0.8,
    point.padding = 0.5,
    force = 8,
    seed = 123
  ) +
  
  # Customize appearance
  scale_shape_manual(values = c("Hill" = 17, "Valley" = 16, "Split" = 15)) +
  scale_color_manual(values = c(
    "Matches Theory" = "gray70", 
    "Misclassified (Low Complexity Hill)" = "#3366CC",
    "Misclassified (High Complexity Valley)" = "#339900"
  )) +
  scale_size_manual(values = c(
    "Matches Theory" = 2, 
    "Misclassified (Low Complexity Hill)" = 3.5,
    "Misclassified (High Complexity Valley)" = 3.5
  )) +
  
  # Add labels
  labs(
    title = "Cell Complexity vs. Case Marking in Hill/Valley Languages",
    x = "Cell Complexity (scaled)",
    y = "Case Marking Complexity (scaled)",
    shape = "Classification",
    color = "Theory Alignment",
    size = "Theory Alignment"
  ) +
  
  # Visual theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```

